The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-07 rational.tex
intro

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclclclclcl}
a &=& a + 0 &=& a + ((-a) + a) \\
  &=& (a + (-a)) + a &=& 0 + a &=& a
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation.  On the left we have a producer that can supply a value of
type $a$ and on the right we have a consumer that demands a value of type
$a$. For the purposes of the example, think of the producer as someone having
\$20.00 and the consumer as someone demanding \$20.00. It is certainly
possible to directly connect the producer and consumer using the algebraic
identity $a=a$: such a connection would be computationally modeled by the
identity function as expected. However, the algebraic proof above implements
this connection in a much more interesting way: the transformation $0 = (-a)
+ a$ is used at the consumer site to create money and a corresponding
debt. The consumer gets the money immediately and the debt travels
``backwards'' towards the producer where it is paid.  In other words, this
computation models something similar to a credit card transaction. We discuss
this example in detail in Sec.~\ref{sub:neg} and only focus on two important
points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
  constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless in its
  treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the resources (money) from the
  propagation of the services. Not every transaction needs both the resources
  and services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as the
  infrastructure can be trusted with such promises. Computationally, this
  means that negative types allow us to break the dependency between
  producers and consumers and let each compute independently.
\end{itemize}

>>>>

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclll}
&& a & \mbox{initial~$a$} & (0) \\
&=& a + 0 & \mbox{$0$~is~identity~for~$+$} & (1) \\
&=& a + (-a + a) & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (2) \\
&=& (a + (-a)) + a & \mbox{$+$~is~associative} & (3) \\
&=& 0 + a & \mbox{$-a$~is~the~additive~inverse~of~$a$} & (4) \\
&=& a & \mbox{$0$~is~identity~for~$+$} & (5)
\end{array}\]
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism with a non-trivial and interesting computational
interpretation. Consider the starting $a$ as being supplied by a producer of
values and the final $a$ as being demanded by a consumer. For the purposes of
the example, think of the producer as someone having \$20.00 and the consumer
as someone demanding \$20.00. It is certainly possible to directly connect
the producer and consumer using the algebraic identity $a=a$: such a
connection would be computationally modeled by the identity function as
expected. However, the algebraic proof above implements this connection in a
much more interesting way. As Sec.~\ref{sec:neg} explains in detail, the
flow of the computation is as follows:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type $a+0$; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), the tag on the \$20.00 changes to indicate that it is
  in the left-left summand, i.e., the value is now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}

Collecting the various events that happen at the same lines, we can
reorganize the above computation as follows:
\begin{itemize}
\item At line 0, we have the value {{20}} flowing forward;
\item At line 1, we have the value {{left 20}} flowing forward;
\item At line 2, we have the value {{left 20}} flowing forward or the value
  {{right (left 20)}} flowing backwards or the value {{right (right 20)}}
  flowing forward;
\item At line 3, we have the value {{left (left 20)}} flowing forward or the
  value {{left (right 20)}} flowing backward or the value {{right 20}}
  flowing forward;
\item At line 4, we have the value {{right 20}} flowing forward;
\item At line 5, we have the value {{20}} flowing forward.
\end{itemize}

Use the above to argue that we have something like a credit card transaction!

the transformation $0 = (-a) + a$ is used at the
consumer site to create money and a corresponding debt. The consumer gets the
money immediately and the debt travels ``backwards'' towards the producer
where it is paid.  In other words, this computation models something similar
to a credit card transaction. We discuss this example in detail in
Sec.~\ref{sub:neg} and only focus on two important points here:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
  constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless in its
  treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the resources (money) from the
  propagation of the services. Not every transaction needs both the resources
  and services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as the
  infrastructure can be trusted with such promises. Computationally, this
  means that negative types allow us to break the dependency between
  producers and consumers and let each compute independently.
\end{itemize}
